---
title: "The Impact of Aging, Dementia, and SES on Cognitive Decline"
subtitle: "A Longitudinal Study Using Linear Mixed Models"
author: "Adasia M., Bess T., Preethi R."
execute:
  echo: true
  warning: false
  message: false
  error: false
format: 
  revealjs:
    theme: beige
    transition: fade
    embed-resources: true
    slide-number: false
    width: 1600
    height: 900
    df-print: paged
    html-math-method: katex
editor: source
pdf-separate-fragments: true
fig-align: center
---

## Introduction

Linear Mixed Models (LMMs) are powerful statistical tools designed to analyze data with complex structures, such as hierarchical data (e.g., individuals within groups) or repeated measures (e.g., assessments taken over time). Unlike traditional methods, LMMs account for variations at both the group level and the individual level, making them ideal for studying patterns of cognitive decline across aging populations.

### Why LMMs?

-   **Fixed Effects**: Capture overall population trends, like how age or socioeconomic status influences cognitive function on average.
-   **Random Effects**: Model differences between individuals, accounting for unique trajectories over time or variability across participants.
-   **Data Flexibility**: Handle missing data and unbalanced datasets effectively, ensuring reliable results even when some observations are incomplete.

## Literature Review

### Linear Mixed Models: An Overview

-   Extend simple linear regression by incorporating **fixed effects** (population-level) and **random effects** (subject-level variability).
-   Ideal for analyzing longitudinal and hierarchical datasets.
-   Provide unbiased estimates even with missing data (Bates, 2014; Gelman & Hill, 2007).

### Covariance Structure

-   Essential for modeling dependencies between repeated measures (Starkweather, 2010).
-   Positive covariance: variables move together.
-   Negative covariance: variables move in opposite directions.
-   Allows explicit modeling of within-subject variability, critical for longitudinal studies.

------------------------------------------------------------------------

## Literature Review Continued

### Robust Estimation

-   Methods that minimize the influence of outliers, ensuring parameter reliability (Agostinelli & Yohai, 2016).
-   Non-robust techniques like OLS regression are highly sensitive to extreme values and may yield biased results.

### Challenges of Traditional Methods

-   Traditional linear models assume **independence of observations**, often violated in clustered data (Barr et al., 2013).

-   Handling missing data through listwise deletion or imputation can introduce bias (Enders, 2010).

------------------------------------------------------------------------

## Methods

### Our Focus

This project applies LMMs to investigate:

-   The influence of **age**, **dementia status**, and **socioeconomic factors** on cognitive decline.

-   Cognitive function measured using the **Mini-Mental State Examination (MMSE)**, a widely-used tool for assessing cognitive impairment.

-   How repeated assessments over time help us understand individual changes in cognitive function while capturing broader population trends.

## Methods

### Dataset: OASIS-Longitudinal MRI Data in Nondemented and Demented Older Adults

-   Source: [NCBI](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2895005/)

**Summary of the chosen dataset:**

-   Includes MRI scans of 150 subjects aged 60-96.
-   Longitudinal data collected over 373 sessions.
-   Subjects scanned at least twice, with visits separated by at least a year.
-   Clinical Dementia Rating (CDR) categorized as nondemented or with mild Alzheimer’s disease.
-   72 were nondemented throughout, while 64 were demented initially and remained so.
-   Participants were right-handed, consisting of 62 men and 88 women.
-   Each session had 3-4 T1-weighted MRI scans.

## Methods

### Modeling Approach

-   **Outcome Variable**: Mini-Mental State Exam (MMSE) score.

-   **Fixed Effects**: Age, dementia status, SES, and interactions.

-   **Random Effects**: Subject ID to capture individual variability.

-   **Software**: R (packages: `lme4`, `tableone` for data summaries).

------------------------------------------------------------------------

## R script for Table One {.smaller}

```{r, warning=FALSE, echo=T, message=FALSE, result ='hide'}

#install.packages("tableone")
library(tableone)

#install.packages("readr")
library(readr)

oasis_longitudinal <- read_csv("C:/IDC6940/IDC6940_BDP/oasis_longitudinal.csv")

# Load necessary library
library(tableone)

# Recode M/F for easier handling in Table 1 (optional)
oasis_longitudinal$Gender <- ifelse(oasis_longitudinal$'M/F' == "M", "Male", "Female")

# Define the variables to be included in Table 1
vars <- c("Age", "Gender", "Hand", "EDUC", "SES", "MMSE", "CDR", "eTIV", "nWBV", "ASF")

# Define the stratifying variable (Group)
strata <- "Group"

# Create Table One
table1 <- CreateTableOne(vars = vars, strata = strata, data = oasis_longitudinal, factorVars = c("Gender", "Hand", "CDR"))

# Print the table with p-values
print(table1, showAllLevels = TRUE, smd = TRUE)

```

## Table One {.bigger}

```{=html}
<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-w7ak{border-color:#ffffff;font-family:Georgia, serif !important;text-align:left;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-w7ak"></th>
    <th class="tg-w7ak">level</th>
    <th class="tg-w7ak">Converted</th>
    <th class="tg-w7ak">Demented</th>
    <th class="tg-w7ak">Nondemented</th>
    <th class="tg-w7ak">p</th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-w7ak">n</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">37</td>
    <td class="tg-w7ak">146</td>
    <td class="tg-w7ak">190</td>
    <td class="tg-w7ak"></td>
  </tr>
  <tr>
    <td class="tg-w7ak">Female</td>
    <td class="tg-w7ak">F</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;24 ( 64.9) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;60 ( 41.1) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;129 ( 67.9) </td>
    <td class="tg-w7ak">&lt;0.001</td>
  </tr>
  <tr>
    <td class="tg-w7ak">Male</td>
    <td class="tg-w7ak">M</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;13 ( 35.1) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;86 ( 58.9) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;61 ( 32.1) </td>
    <td class="tg-w7ak"></td>
  </tr>
  <tr>
    <td class="tg-w7ak">Dom Hand</td>
    <td class="tg-w7ak">R</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;37 (100.0) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;146 (100.0) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;190 (100.0) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;NA</td>
  </tr>
  <tr>
    <td class="tg-w7ak">CDR</td>
    <td class="tg-w7ak">0</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;18 ( 48.6) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;0 (&nbsp;&nbsp;0.0) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;188 ( 98.9) </td>
    <td class="tg-w7ak">&lt;0.001</td>
  </tr>
  <tr>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">0.5</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;19 ( 51.4) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;102 ( 69.9) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;2 (&nbsp;&nbsp;1.1) </td>
    <td class="tg-w7ak"></td>
  </tr>
  <tr>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">1</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;0 (&nbsp;&nbsp;0.0) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;41 ( 28.1) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;0 (&nbsp;&nbsp;0.0) </td>
    <td class="tg-w7ak"></td>
  </tr>
  <tr>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">2</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;0 (&nbsp;&nbsp;0.0) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;3 (&nbsp;&nbsp;2.1) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;0 (&nbsp;&nbsp;0.0) </td>
    <td class="tg-w7ak"></td>
  </tr>
  <tr>
    <td class="tg-w7ak">Age</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">&nbsp;&nbsp;79.76 (7.43)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;76.26 (6.94)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;77.06 (8.10)</td>
    <td class="tg-w7ak">0.045</td>
  </tr>
  <tr>
    <td class="tg-w7ak">Education</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">&nbsp;&nbsp;15.46 (2.52)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;13.67 (2.90)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;15.14 (2.74)</td>
    <td class="tg-w7ak">&lt;0.001</td>
  </tr>
  <tr>
    <td class="tg-w7ak">SES</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;1.73 (0.96)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;2.77 (1.20)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;2.39 (1.05)</td>
    <td class="tg-w7ak">&lt;0.001</td>
  </tr>
  <tr>
    <td class="tg-w7ak">MMSE</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">&nbsp;&nbsp;28.68 (1.56)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;24.51 (4.50)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;29.23 (0.88)</td>
    <td class="tg-w7ak">&lt;0.001</td>
  </tr>
  <tr>
    <td class="tg-w7ak">eTIV</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">1459.27 (135.43)</td>
    <td class="tg-w7ak">1485.85 (173.77)</td>
    <td class="tg-w7ak">1495.50 (184.89)</td>
    <td class="tg-w7ak">0.51</td>
  </tr>
  <tr>
    <td class="tg-w7ak">nWBV</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;0.72 (0.04)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;0.72 (0.03)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;0.74 (0.04)</td>
    <td class="tg-w7ak">&lt;0.001</td>
  </tr>
  <tr>
    <td class="tg-w7ak">ASF</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;1.21 (0.11)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;1.20 (0.14)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;1.19 (0.14)</td>
    <td class="tg-w7ak">0.683</td>
  </tr>
</tbody></table>
```
## Data Visualizations {.bigger}

### Age of Participants

```{r, warning=FALSE, echo=FALSE, message=FALSE}
# Load required libraries
library(readr)

# Load the dataset (ensure the correct file path is provided)
oasis_longitudinal <- read_csv("C:/IDC6940/IDC6940_BDP/oasis_longitudinal.csv")

# Visualization: Age of Participants
counts <- table(oasis_longitudinal$Age)
barplot(counts, 
        main = "Age of Participants", 
        xlab = "Age", 
        col = "lightblue", 
        border = "white")
```

## Data Visualizations {.tabset}

### Gender of Participants

```{r, warning=FALSE, echo=FALSE, message=FALSE}
# Load required libraries
library(readr)

# Load the dataset (ensure the correct file path is provided)
oasis_longitudinal <- read_csv("C:/IDC6940/IDC6940_BDP/oasis_longitudinal.csv")
# Visualization: Gender of Participants
counts <- table(oasis_longitudinal$'M/F')
barplot(counts, 
        main = "Gender of Participants", 
        xlab = "Gender", 
        col = "lightgreen", 
        border = "white")

```

## Data Visualizations {.tabset}

### Socioeconomic Status of Participants

```{r, warning=FALSE, echo=FALSE, message=FALSE}
# Load required libraries
library(readr)

# Load the dataset (ensure the correct file path is provided)
oasis_longitudinal <- read_csv("C:/IDC6940/IDC6940_BDP/oasis_longitudinal.csv")
# Visualization: Socioeconomic Status
counts <- table(oasis_longitudinal$SES)
barplot(counts, 
        main = "Socioeconomic Status of Participants", 
        xlab = "SES", 
        col = "lightcoral", 
        border = "white")

```

## Data Visualizations {.tabset}

### Normalized Whole Brain Volume vs Age

```{r, warning=FALSE, echo=FALSE, message=FALSE}
# Load required libraries
library(readr)

# Load the dataset (ensure the correct file path is provided)
oasis_longitudinal <- read_csv("C:/IDC6940/IDC6940_BDP/oasis_longitudinal.csv")
# Visualization: nWBV vs Age
x <- oasis_longitudinal$nWBV
y <- oasis_longitudinal$Age
plot(x, y, 
     main = "Normalized Whole Brain Volume vs Age", 
     xlab = "nWBV", 
     ylab = "Age", 
     pch = 19, 
     col = "purple", 
     frame = FALSE)
abline(lm(y ~ x, data = oasis_longitudinal), col = "blue", lwd = 2)

```

# Age as a Predictor of Cognitive Decline

By: Adasia McClinton

## Data Cleaning and Ingestion

```{r}
data <- read.csv("C:/IDC6940/IDC6940_BDP/oasis_longitudinal.csv")
library(mice)
imputed_data <- mice(data, m = 5, method = 'pmm')
imputed_list <- lapply(1:5, function(i) complete(imputed_data, i))
```

## Model 0: Age as the Sole Predictor

-   Assess whether **age** alone is a significant predictor of cognitive decline as measured by MMSE scores.

$$
\text{MMSE}_{ij} = 29.660 - 0.031 \times \text{Age}_{ij} + u_{0j} + \epsilon_{ij}
$$

-   **Intercept**: 29.660 (baseline MMSE score).
-   **Age Coefficient**: -0.031 (small, non-significant decline with age).

### Key Findings

-   Age alone is not a significant predictor of MMSE scores.
-   Indicates additional variables may explain cognitive decline.

## Model 0: Age as the Sole Predictor

### R Code

```{r}

# Model 0: Age as the sole predictor
library("lme4")
library("mitml")
library("Matrix")
model0 <- lapply(imputed_list, function(data) lmer(MMSE ~ Age + (1 | Subject.ID), data = data))
pooled_results <- testEstimates(model0, method = "D2")
summary(pooled_results)

```

-   **Results**: The age effect on MMSE was small and non-significant, suggesting that age alone does not fully explain cognitive decline.

## Model 1: Adding Dementia Severity and SES

-   Examine the combined effect of age, dementia severity (CDR), and SES

    $$\text{MMSE}_{ij} = 34.544 - 0.070 \times \text{Age}_{ij} - 4.699 \times \text{GD}_{ij} - 0.231 \times \text{SES}_{ij} + 0.008 \times (\text{Age} \times \text{GD})_{ij} + u_{0j} + \epsilon_{ij}$$

```{=html}
<!-- -->
```
-   **Age Coefficient**: -0.070 (non-significant decline).

-   **CDR Coefficient**: -4.699 (significant decline in MMSE).

-   **SES Coefficient**: -0.231 (small decline).

### Key Findings

-   Dementia severity significantly predicts MMSE scores.

-   Age and SES are not significant predictors in this model.

## Model 1: Adding Dementia Severity and SES

### R Code

```{r}
model1 <- lapply(imputed_list, function(data) {
  lmer(MMSE ~ Age * Group + SES + (1 | Subject.ID), data = data)
})
pooled_results <- testEstimates(model1, method = "D2")
summary(pooled_results)
```

-   **Results**: This model showed no significant effects for age, SES, or age-by-group interactions, indicating that additional factors may better explain cognitive decline.

## Final Model: Significant Predictors Only

Focus on significant predictors: **CDR** and **nWBV**.

$$
\text{MMSE}_{ij} = 17.894 - 5.441 \times \text{CDR}_{ij} + 15.038 \times \text{nWBV}_{ij} + u_{0j} + \epsilon_{ij}
$$

-   **CDR Coefficient**: -5.441 (strong negative effect).

-   **nWBV Coefficient**: 15.038 (positive effect).

### Key Findings

-   CDR strongly predicts cognitive decline.

-   Higher nWBV mitigates decline.

## Final Model: Significant Predictors Only

### R Code

```{r}
final_model <- lapply(imputed_list, function(data) {
  lmer(MMSE ~ CDR + nWBV + (1 | Subject.ID), data = data)
})
pooled_final_results <- testEstimates(final_model, method = "D2")
summary(pooled_final_results)
```

-   **Results**: CDR was significantly associated with MMSE scores, showing that higher dementia severity leads to lower cognitive function. Additionally, higher nWBV was associated with higher MMSE, supporting its role in cognitive preservation.

## Conclusions

### Key Takeaways

-   Age alone does not predict cognitive decline.

-   Dementia severity (CDR) is a strong negative predictor.

-   Higher brain volumes (nWBV) are protective against cognitive decline.

## References

-   Agostinelli, C., & Yohai, V. J. (2016). Composite robust estimators for linear mixed models. *Journal of the American Statistical Association*, 111(516), 1764-1774. <https://doi.org/10.1080/01621459.2015.1115358>

-   Bates, D. (2014). Fitting linear mixed-effects models using lme4. arXiv preprint arXiv:1406.5823.

-   Baayen, R. H., Davidson, D. J., & Bates, D. M. (2008). Mixed-effects modeling with crossed random effects for subjects and items. *Journal of Memory and Language*, 59(4), 390-412.

-   Barr, D. J., Levy, R., Scheepers, C., & Tily, H. J. (2013). Random effects structure for confirmatory hypothesis testing: Keep it maximal. *Journal of Memory and Language*, 68(3), 255-278.

-   Bruin, J. (2006). newtest: command to compute new test. UCLA: Statistical Consulting Group. <https://stats.oarc.ucla.edu/stata/ado/analysis/>.

## References (continued)

-   Enders, C. K. (2010). *Applied Missing Data Analysis*. Guilford Press.

-   Gelman, A., & Hill, J. (2007). *Data Analysis Using Regression and Multilevel/Hierarchical Models*. Cambridge University Press.

-   Starkweather, J. (2010). Linear mixed effects modeling using R. Unpublished Manuscript.

## 
