---
title: "Dataset Summaries"
format: html
editor: visual
self_contained: true
---

### Dataset Summary: Mental Health in Tech Survey

-   **Overview:**

    -   This dataset is derived from a 2014 survey measuring attitudes toward mental health and the prevalence of mental health disorders in the technology workplace.

    -   Encourages further analysis with data from an ongoing 2016 survey.

-   **Content:**

    -   Includes responses on various aspects related to mental health in the workplace, such as:

        -   **Timestamp:** Date and time of the survey response.

        -   **Demographics:** Age, gender, country, and state (for U.S. respondents).

        -   **Employment Details:** Self-employment status, number of employees at the organization, remote work status, and whether the employer is a tech company.

        -   **Mental Health Treatment:** Family history of mental illness, treatment-seeking behavior, and perceived interference of mental health conditions with work.

        -   **Employer Policies:** Availability of mental health benefits, care options, wellness programs, and resources for seeking help.

        -   **Anonymity and Leave:** Protection of anonymity in treatment resources and ease of taking medical leave for mental health issues.

        -   **Consequences and Support:** Perceived negative consequences of discussing mental health with employers or coworkers, willingness to discuss mental health issues, and observations of consequences for others.

        -   **Comments:** Additional notes from respondents.

-   **Purpose:**

    -   To analyze the mental health landscape in the tech industry and inform policies and practices that support mental health awareness and treatment in the workplace.

```{r}
library(readr)
survey <- read_csv("survey.csv")
head(survey)
```

### Dataset Summary: MRI and Alzheimer's (Longitudinal Data)

-   **Overview:**

    -   The dataset is part of the Open Access Series of Imaging Studies (OASIS), aimed at providing MRI brain data to the scientific community to advance neuroscience research.

-   **Content:**

    -   **Longitudinal MRI Data:**

        -   Comprises 150 subjects aged 60 to 96 with scans taken on two or more occasions, spaced at least one year apart (totaling 373 imaging sessions).

        -   Subjects include both men and women, all right-handed.

        -   72 subjects remained nondemented throughout the study.

        -   64 subjects were characterized as demented at the initial visit and remained so for subsequent scans, including 51 individuals with mild to moderate Alzheimer’s disease.

        -   14 subjects transitioned from nondemented to demented during the study.

-   **Variables:**

    -   Key attributes include Subject ID, MRI ID, group classification (demented/nondemented), visit number, age, education level (EDUC), socioeconomic status (SES), Mini-Mental State Examination (MMSE) scores, Clinical Dementia Rating (CDR), estimated total intracranial volume (eTIV), normalized whole brain volume (nWBV), and atrophy scaling factor (ASF).

-   **Research Implications:**

    -   The dataset is intended to facilitate research into dementia and Alzheimer’s disease, particularly in predictive modeling and understanding disease progression.

-   **Acknowledgements:**

    -   Researchers are encouraged to acknowledge specific grant numbers when publishing findings based on this dataset.

```{r}
library(readr)
oasis_longitudinal <- read_csv("oasis_longitudinal.csv")
head(oasis_longitudinal)
```
